Location and Scale-Invariant Power Transformations for Transforming Data to Normality

### 2025-10-29 manuscript_rmarkdown/manuscript.tex

\title{Location and Scale-Invariant Power Transformations for
Transforming Data to Normality}
\author{Alex Zwanenburg, Steffen Löck}
\date{2025-10-28}

@@

Each sequence was created as follows: first \(n\) values are sampled
from \(\mathcal{N}(0, 1)\). Then, the elements outside the central
portion of the sequence
(i.e.~\(\mathbf{X} \setminus \mathbf{X}_{\text{c}}\)) are replaced.
Elements with \(pi < \frac{1-\kappa}{2}\) are replaced with values
sampled from \(\mathcal{U}(-10,\min(\mathbf{X}_{\text{c}}))\), and
elements with \(p_i > \frac{1 + \kappa}{2}\) are replaced with values
sampled from \(\mathcal{U}(\max(\mathbf{X}_{\text{c}}), 10)\).

@@

\caption{Observed type I error rates for central normality tests in the presence of data sampled from central normal distributions with $\kappa \in \left\{0.6, 0.8, 1.0\right\}$. Three empirical central normality (ECN) tests were performed, parameterised with same value for the central portion $\kappa$. Additionally, the Shapiro-Wilk test was assessed. ECN: empirical central normality}\label{fig:empirical-central-normality-error_rate}
\end{figure}

@@

\caption{Bi-modal distributions and empirical central normality test results. The feature (black) is a mixture of two identical sample sets (blue and orange, $n = 100$) drawn from normal distributions that are offset by a distance $d$. The empirical central normality ($\kappa = 0.80$) is used to test the hypothesis that the mixture is sampled from a central normal distribution. As may be observed, with increasing offset $d$ the test's p-value decreases. Quantile-quantile plots are drawn below each distribution.}\label{fig:empirical-central-normality-examples}
\end{figure}

@@

A common feature in health-related datasets is age. Here we use data on
228 patients with lung cancer that was collected and published by
Loprinzi et al. \citep{Loprinzi1994-cd}. The age in the cohort was
\(62.4 \pm 9.1\) (mean ± standard deviation) years. Applying
conventional and invariant Yeo-Johnson transformations to patient age
yielded the following results, see Figure
\ref{fig:experimental-results-invariance}: no transformation
(\(\sum r_i = 16.5\), \(p = 0.69\)); conventional transformation
(\(\lambda = 2.0\), \(\sum r_i = 11.5\), \(\mu_{YJ} = 1.8 \cdot 10^3\),
\(\sigma_{YJ} = 0.5 \cdot 10^3\), \(p = 0.96\)); Raymaekers and
Rousseeuw's robust adaptation (\(\lambda = 2.0\), \(\sum r_i = 11.5\),
\(\mu_{YJ} = 1.8 \cdot 10^3\), \(\sigma_{YJ} = 0.5 \cdot 10^3\),
\(p = 0.96\)); location- and scale-invariant transformation
(\(\lambda = 1.3\), \(\sum r_i = 8.8\), \(\mu_{YJ} = -1.2\),
\(\sigma_{YJ} = 1.1\), \(p = 0.98\)); and robust location- and
scale-invariant transformation (\(\lambda = 1.3\), \(\sum r_i = 9.3\),
\(\mu_{YJ} = -1.0\), \(\sigma_{YJ} = 1.1\), \(p = 0.93\)).

Location- and scale-invariant transformation led to a lower overall
residual error, indicating a better transformation to normality.
Conventional transformations inflated the mean \(\mu_{YJ}\) and standard
deviation \(\sigma_{YJ}\) of the age feature after transformation. The
empirical central normality test did not detect any statistically
significant deviations from central normality for any transformation
(all \(p \geq 0.93\)).

@@

Gorman, Williams and Fraser recorded body mass (in grams) of 342
penguins of three different species \citep{Gorman2014-eo}. The body mass
was \((4.2 \pm 0.8) \cdot 10^3\) (mean ± standard deviation) grams, and
not centrally normal (\(p < 0.001\)). Applying conventional and
invariant Yeo-Johnson transformations to body mass yielded the following
results, see Figure \ref{fig:experimental-results-invariance}: no
transformation (residual sum \(\sum r_i = 48.0\), \(p < 0.001\));
conventional transformation (\(\lambda = -0.5\), \(\sum r_i = 32.2\),
\(\mu_{YJ} = 2.1\), \(\sigma_{YJ} = 4 \cdot 10^{-3}\), \(p = 0.10\));
Raymaekers and Rousseeuw's robust adaptation (\(\lambda = -0.5\),
\(\sum r_i = 32.2\), \(\mu_{YJ} = 2.1\),
\(\sigma_{YJ} = 4 \cdot 10^{-3}\), \(p = 0.10\)); location- and
scale-invariant transformation (\(\lambda = 0.5\), \(\sum r_i = 26.8\),
\(\mu_{YJ} = 0.9\), \(\sigma_{YJ} = 0.9\), \(p = 0.28\)); and robust
location- and scale-invariant transformation (\(\lambda = 0.3\),
\(\sum r_i = 22.0\), \(\mu_{YJ} = 0.7\), \(\sigma_{YJ} = 0.9\),
\(p = 0.69\)).

Location- and scale-invariant transformation produced a lower overall
residual errors, indicating a better transformation. Moreover,
conventional transformations led to low standard deviation
\(\sigma_{YJ}\) of the body mass feature after transformation. The
empirical central normality test did not detect any statistically
significant deviations from central normality for any transformation
(all \(p \geq 0.10\)).

@@

Every transformation reduced the residual sum. None of the
transformations yielded a centrally normal distribution. Conventional
transformations had high values for the \(\lambda\) parameter, which
could lead to numerical issues.

@@

The Top Gear dataset contains data on 297 vehicles that appeared on the
BBC television show \emph{Top Gear} \citep{Alfons2021-kc}. Within this
dataset, the fuel consumption feature contains outliers due to highly
fuel-efficient vehicles. Applying conventional and invariant Yeo-Johnson
transformations to the fuel consumption feature yielded the following
results, see Figure \ref{fig:experimental-results-outlier-robustness}:
no transformation (residual sum \(\sum r_i = 54\), \(p=0.72\));
conventional transformation (\(\lambda = -0.1\), \(\sum r_i = 55\),
\(\mu_{YJ} = 3.0\), \(\sigma_{YJ} = 0.3\), \(p < 0.001\)); Raymaekers
and Rousseeuw's robust adaptation (\(\lambda = 0.8\), \(\sum r_i = 48\),
\(\mu_{YJ} = 29\), \(\sigma_{YJ} = 15\), \(p=0.37\)); location- and
scale-invariant transformation (\(\lambda = -1.3\), \(\sum r_i = 44\),
\(\mu_{YJ} = 0.5\), \(\sigma_{YJ} = 0.1\), \(p < 0.001\)); and robust
location- and scale-invariant transformation (\(\lambda = 1.0\),
\(\sum r_i = 56\), \(\mu_{YJ} = 1.7\), \(\sigma_{YJ} = 1.0\),
\(p=0.76\)).

Outliers cause non-robust transformations to fail to transform the data
to a centrally normal distribution (empirical central normality test
\(p < 0.001\) for conventional and invariant transformations). Robust
transformations produce distributions that are centrally normal
(empirical central normality test \(p > 0.05\)).

@@

The ischemic stroke dataset contains historic data from 126 patients
with risk at ischemic stroke \citep{Kuhn2019-kt}. These patients
underwent Computed Tomography Angiography to characterize the carotid
artery blockages. Angiography imaging was then assessed, and various
characteristics related to the blood vessels and the disease are
measured. The maximum arterial wall thickness feature contains several
instances with outlier values. Applying conventional and invariant
Yeo-Johnson transformations to this feature yielded the following
results, see Figure \ref{fig:experimental-results-outlier-robustness}:
no transformation (residual sum \(\sum r_i = 110\), \(p=0.83\));
conventional transformation (\(\lambda = -0.7\), \(\sum r_i = 30\),
\(\mu_{YJ} = 1.0\), \(\sigma_{YJ} = 0.1\), \(p=0.003\)); Raymaekers and
Rousseeuw's robust adaptation (\(\lambda = 1.1\), \(\sum r_i = 136\),
\(\mu_{YJ} = 7.2\), \(\sigma_{YJ} = 14.3\), \(p=0.88\)); location- and
scale-invariant transformation (\(\lambda = 0.2\), \(\sum r_i = 12\),
\(\mu_{YJ} = -11.8\), \(\sigma_{YJ} = 6.9\), \(p=0.15\)); and robust
location- and scale-invariant transformation (\(\lambda = -0.3\),
\(\sum r_i = 30\), \(\mu_{YJ} = 0.8\), \(\sigma_{YJ} = 0.2\),
\(p=0.18\)).

The conventional non-robust transformation failed to produce a centrally
normal distribution (empirical central normality test \(p=0.003\)).
Robust transformations produce distributions that are centrally normal
(empirical central normality test \(p > 0.05\)).

>>>>

\title{Location and Scale-Invariant Power Transformations for
Transforming Data to Normality}
\author{Alex Zwanenburg, Steffen Löck}
\date{2025-10-29}

@@

Each sequence was created as follows: first \(n\) values are sampled
from \(\mathcal{N}(0, 1)\). Then, the elements outside the central
portion of the sequence
(i.e.~\(\mathbf{X} \setminus \mathbf{X}_{\text{c}}\)) are replaced.
Elements with \(p_i < \frac{1-\kappa}{2}\) are replaced with values
sampled from \(\mathcal{U}(-10,\min(\mathbf{X}_{\text{c}}))\), and
elements with \(p_i > \frac{1 + \kappa}{2}\) are replaced with values
sampled from \(\mathcal{U}(\max(\mathbf{X}_{\text{c}}), 10)\).

@@

\caption{Observed type I error rates for central normality tests in the presence of data sampled from central normal distributions with $\kappa \in \left\{0.6, 0.8, 1.0\right\}$. Three empirical central normality (ECN) tests were performed, parameterised with same value for the central portion $\kappa$. Additionally, the Shapiro-Wilk test was assessed. The dashed line indicates the expected type I error rate for rejecting the null hypothesis at significance level $\alpha = 0.05$. ECN: empirical central normality}\label{fig:empirical-central-normality-error_rate}
\end{figure}

@@

\caption{Bi-modal distributions and empirical central normality test results. The feature (black) is a mixture of two identical sample sets (blue and orange, $n = 100$) drawn from normal distributions that are offset by a distance $d$. The empirical central normality test ($\kappa = 0.80$) is used to test the hypothesis that the mixture is sampled from a central normal distribution. As may be observed, with increasing offset $d$ the test's p-value decreases. Quantile-quantile plots are drawn below each distribution.}\label{fig:empirical-central-normality-examples}
\end{figure}

@@

A common feature in health-related datasets is age. Here we use data on
228 patients with lung cancer that was collected and published by
Loprinzi et al. \citep{Loprinzi1994-cd}. The age in the cohort was
\(62.4 \pm 9.1\) (mean ± standard deviation) years. Applying
conventional and invariant Yeo-Johnson transformations to patient age
yielded the following results, see Figure
\ref{fig:experimental-results-invariance}: no transformation
(\(\sum r_i = 16.5\), \(p = 0.54\)); conventional transformation
(\(\lambda = 2.0\), \(\sum r_i = 11.5\), \(\mu_{YJ} = 1.8 \cdot 10^3\),
\(\sigma_{YJ} = 0.5 \cdot 10^3\), \(p = 0.92\)); Raymaekers and
Rousseeuw's robust adaptation (\(\lambda = 2.0\), \(\sum r_i = 11.5\),
\(\mu_{YJ} = 1.8 \cdot 10^3\), \(\sigma_{YJ} = 0.5 \cdot 10^3\),
\(p = 0.92\)); location- and scale-invariant transformation
(\(\lambda = 1.3\), \(\sum r_i = 8.8\), \(\mu_{YJ} = -1.2\),
\(\sigma_{YJ} = 1.1\), \(p = 0.95\)); and robust location- and
scale-invariant transformation (\(\lambda = 1.3\), \(\sum r_i = 9.3\),
\(\mu_{YJ} = -1.0\), \(\sigma_{YJ} = 1.1\), \(p = 0.87\)).

Location- and scale-invariant transformation led to a lower overall
residual error, indicating a better transformation to normality.
Conventional transformations inflated the mean \(\mu_{YJ}\) and standard
deviation \(\sigma_{YJ}\) of the age feature after transformation. The
empirical central normality test did not detect any statistically
significant deviations from central normality for any transformation
(all \(p \geq 0.87\)).

@@

Gorman, Williams and Fraser recorded body mass (in grams) of 342
penguins of three different species \citep{Gorman2014-eo}. The body mass
was \((4.2 \pm 0.8) \cdot 10^3\) (mean ± standard deviation) grams, and
not centrally normal (\(p < 0.001\)). Applying conventional and
invariant Yeo-Johnson transformations to body mass yielded the following
results, see Figure \ref{fig:experimental-results-invariance}: no
transformation (residual sum \(\sum r_i = 48.0\), \(p < 0.001\));
conventional transformation (\(\lambda = -0.5\), \(\sum r_i = 32.2\),
\(\mu_{YJ} = 2.1\), \(\sigma_{YJ} = 4 \cdot 10^{-3}\), \(p = 0.03\));
Raymaekers and Rousseeuw's robust adaptation (\(\lambda = -0.5\),
\(\sum r_i = 32.2\), \(\mu_{YJ} = 2.1\),
\(\sigma_{YJ} = 4 \cdot 10^{-3}\), \(p = 0.03\)); location- and
scale-invariant transformation (\(\lambda = 0.5\), \(\sum r_i = 26.8\),
\(\mu_{YJ} = 0.9\), \(\sigma_{YJ} = 0.9\), \(p = 0.12\)); and robust
location- and scale-invariant transformation (\(\lambda = 0.3\),
\(\sum r_i = 22.0\), \(\mu_{YJ} = 0.7\), \(\sigma_{YJ} = 0.9\),
\(p = 0.48\)).

Location- and scale-invariant transformation yielded lower overall
residual errors, indicating a better transformation to central
normality, which is reflected by the empirical central normality test
results. Conventional transformations led to low standard deviation
\(\sigma_{YJ}\) of the body mass feature after transformation, and the
null hypothesis of the empirical central normality test was rejected (at
significance level \(\alpha = 0.05\)) in both cases.

@@

Every transformation reduced the residual sum compared to the original
data. None of the transformations yielded a centrally normal
distribution. Conventional transformations had high values for the
\(\lambda\) parameter, which could lead to numerical issues.

@@

The Top Gear dataset contains data on 297 vehicles that appeared on the
BBC television show \emph{Top Gear} \citep{Alfons2021-kc}. Within this
dataset, the fuel consumption feature contains outliers due to highly
fuel-efficient vehicles. Applying conventional and invariant Yeo-Johnson
transformations to the fuel consumption feature yielded the following
results, see Figure \ref{fig:experimental-results-outlier-robustness}:
no transformation (residual sum \(\sum r_i = 54\), \(p=0.55\));
conventional transformation (\(\lambda = -0.1\), \(\sum r_i = 55\),
\(\mu_{YJ} = 3.0\), \(\sigma_{YJ} = 0.3\), \(p < 0.001\)); Raymaekers
and Rousseeuw's robust adaptation (\(\lambda = 0.8\), \(\sum r_i = 48\),
\(\mu_{YJ} = 29\), \(\sigma_{YJ} = 15\), \(p=0.21\)); location- and
scale-invariant transformation (\(\lambda = -1.3\), \(\sum r_i = 44\),
\(\mu_{YJ} = 0.5\), \(\sigma_{YJ} = 0.1\), \(p < 0.001\)); and robust
location- and scale-invariant transformation (\(\lambda = 1.0\),
\(\sum r_i = 56\), \(\mu_{YJ} = 1.7\), \(\sigma_{YJ} = 1.0\),
\(p=0.59\)).

Outliers caused non-robust transformations to fail the empirical central
normality test for transforming the feature to a centrally normal
distribution.

@@

The ischemic stroke dataset contains historic data from 126 patients
with risk at ischemic stroke \citep{Kuhn2019-kt}. These patients
underwent Computed Tomography Angiography to characterize the carotid
artery blockages. Angiography imaging was then assessed, and various
characteristics related to the blood vessels and the disease are
measured. The maximum arterial wall thickness feature contains several
instances with outlier values. Applying conventional and invariant
Yeo-Johnson transformations to this feature yielded the following
results, see Figure \ref{fig:experimental-results-outlier-robustness}:
no transformation (residual sum \(\sum r_i = 110\), \(p=0.73\));
conventional transformation (\(\lambda = -0.7\), \(\sum r_i = 30\),
\(\mu_{YJ} = 1.0\), \(\sigma_{YJ} = 0.1\), \(p < 0.001\)); Raymaekers
and Rousseeuw's robust adaptation (\(\lambda = 1.1\),
\(\sum r_i = 136\), \(\mu_{YJ} = 7.2\), \(\sigma_{YJ} = 14.3\),
\(p=0.80\)); location- and scale-invariant transformation
(\(\lambda = 0.2\), \(\sum r_i = 12\), \(\mu_{YJ} = -11.8\),
\(\sigma_{YJ} = 6.9\), \(p=0.07\)); and robust location- and
scale-invariant transformation (\(\lambda = -0.3\), \(\sum r_i = 30\),
\(\mu_{YJ} = 0.8\), \(\sigma_{YJ} = 0.2\), \(p=0.09\)).

For conventional non-robust transformation the null hypothesis was
rejected (at \(\alpha = 0.05\)), indicating that they failed to produce
a transformed feature that followed a central normal distribution.
